---
title: "Shayan_Gerami_IDS472_HW3"
output:
  html_document:
    df_print: paged
  pdf_document: default
date: "2024-10-06"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Show code (set to FALSE if you prefer not to display it)
  results = 'hide',    # Hide results
  message = FALSE,     # Suppress messages
  warning = FALSE,     # Suppress warnings
  error = FALSE        # Suppress errors
)
```

# Install packages, load and import data

```{r Import packages}

# Import libraries
library(dplyr)
library(ggplot2)
library(ISLR)
library(rpart)
library(rpart.plot)
library(rattle)

```

```{r Load data}

# Load rename, attach and view data
df <- read.csv("C:/Users/shaya/Downloads/Placement.csv") # Replace your path
attach(df) # For easier access to data columns

```

# Data Cleaning 
## Covering questions 1-3
- Convert data types if needed
- Replace  missing values
- Handel outliers

```{r Data Prep}
# Check data structure
str(df)

# Convert characters to factor
df <- mutate_if(df, is.character, as.factor)

# Get number of N/As 
sum(is.na(df))

# Nulls are only in salary, not placed records have N/A salary so we'll replace them by 0
df <- df %>% mutate_if(is.numeric, ~ ifelse(is.na(.), 0, .))

```

```{r Replace Outliers}

# Function to detect outliers
outlier_check <- function(x) {
  
  Q1 <- quantile(x, 0.25)
  Q3 <- quantile(x, 0.75)
  
  IQR <-  Q3 - Q1
  
  LB <- Q1 - 1.5*IQR
  UB <- Q3 + 1.5*IQR
  
  return (x < LB | x > UB)
}

# Apply the function to all numeric columns 
outliers <- lapply(df[, sapply(df, is.numeric)], outlier_check)
outliers # See outliers

# Function to replace outliers with median
outlier_replace <- function(x) {
  
  outliers <- outlier_check(x)
  x[outliers] <- median(x)
  
  return(x)
}

# Apply the function to all numeric columns 
df[, sapply(df, is.numeric)] <- lapply(df[, sapply(df, is.numeric)], outlier_replace)

```

# EDA
## Covering questions 4-16

```{r Q4}

# Function to calculate med,mean, min, max for a given variable
min_max_med_mean <- function(x) {
  
  summary <- df %>%
    summarise(
      mean = mean(x, na.rm = TRUE),
      med = median(x, na.rm = TRUE),
      min = min(x, na.rm = TRUE),
      max = max(x, na.rm = TRUE)
    )
  return (summary)
}

#Apply function on desired columns 

# ssc_P
summary_ssc_P <- lapply(ssc_p, min_max_med_mean)
summary_ssc_P # Check

# hsc_P
summary_hsc_p<- lapply(hsc_p, min_max_med_mean)
summary_hsc_p # Check

# degree_p
summary_degree_p<- lapply(degree_p, min_max_med_mean)
summary_degree_p # Check

```

```{r Q5}

# Num students placed and not placed
nrow(filter(df, status == 'Placed'))
nrow(filter(df, status == 'Not Placed'))

```

```{r Q6}

# Avg salary for each gender group
avg_salary_by_gender <- df %>% group_by(gender) %>%
  summarize(avg_salary = mean(salary))

avg_salary_by_gender # Check

```

```{r Q7}

# Create a new var is_placed (1 if placed, 0 if not)
df <- mutate(df, is_placed = if_else(status == "Placed", 1,0))

# Avg of hsc_p by placement status
hsc_P_avg_by_status <- df %>% group_by(is_placed) %>%
  summarize(avg_hcsp = mean(hsc_p))

hsc_P_avg_by_status # Check

```

```{r Q8}

# Avg salary among placed students by specialization
spec_avg_salary_placed <- df %>% filter(is_placed == 1) %>%
  group_by(specialisation) %>%
  summarize(avg_salary = mean(salary)) %>%
  arrange(desc(avg_salary))

spec_avg_salary_placed # Check

```

```{r Q9}

# Percentage of placed among M & F

df_placed <- filter(df, is_placed == 1)
total_placed <- nrow(df_placed)

# For Male
M_percentage <- (nrow(filter(df_placed, gender == "M")) / total_placed) * 100
# For Female
F_percentage <- (nrow(filter(df_placed, gender == "F")) / total_placed) * 100

M_percentage # Check
F_percentage # Check

# Compare placement rates accross genders

placement_rate <- group_by(df, gender) %>%
  summarize(total = n(),
            placed = sum(is_placed == 1),
            placement_rate = (placed / total) * 100)
placement_rate

ggplot(placement_rate, mapping = aes(x=gender, y = placement_rate, fill=gender)) +
  geom_bar(stat = 'identity') +
  labs(title = 'Placement rates among genders' ,
       x = 'Gender',
       y = 'Placement rate')

```

```{r Q10}

# Distribution of placement outcomes
ggplot(df, aes(x = status, fill = status)) +
  geom_bar() +
  labs(title = "Distribution of Placement Outcomes",
       x = "Placement Status",
       y = "Count")

```

```{r Q11}
# Box plot for ssc_p among placed and not placed
ggplot(df, mapping = aes(x = status, y = ssc_p, fill = status)) +
  geom_boxplot() +
  labs(title='ssc_p Boxplot' , 
       x = 'Status')
```

```{r Q12}

# Distreibution of salary among placed 
df_placed <- filter(df, is_placed == "1")

ggplot(df_placed, mapping = aes(x=salary)) +
  geom_histogram() +
  labs(title = 'Distreibution of salary among placed students' ,
       x = 'Salary',
       y = 'Frequency')

```

```{r Q13}

# hsc_p vs degree_p by status
ggplot(df, mapping = aes(x = hsc_p, y = degree_p, color = status)) +
  geom_point() +
  labs(title = 'hsc_p vs degree_p colored by status',
       x = 'hsc_p', y = 'degree_p')


```

```{r Q14}

# Select numericals columns and apply cor()
df_numeric <- select(df, where(is.numeric))
cor(df_numeric, use = 'everything')

```

```{r Q15,16}

# Check potentioal relationship between degree_p and salary based on cor matrix

ggplot(df, mapping = aes(x = degree_p, y = salary)) +
  geom_point() +
  labs(title = 'degree_p vs salary',
       x = 'degree_p', y = 'salary')

ggplot(df, mapping = aes(x = ssc_p, y = is_placed)) +
  geom_point() +
  labs(title = 'ssc_p vs is_placed',
       x = 'ssc_p', y = 'is_placed') #lower ssc_p leads to not getting placed

ggplot(df, mapping = aes(x = hsc_p, y = is_placed)) +
  geom_point() +
  labs(title = 'hsc_p vs is_placed',
       x = 'hsc_p', y = 'is_placed') #lower hsc_p leads to not getting placed
```

# Decision Tree 
## Covering questions 17-24
- Splitting data
- Building model
- Evaluation and pruning

```{r Q17,18}

# Split into train and test
set.seed(421)
index_train <- sample(1:nrow(df), size = 0.7*nrow(df))
train <- df[index_train, ]
test <- df[-index_train, ]

# Remove salary and is_placed from train and test
train <- select(train, -salary, -is_placed)
test <- select(test, -salary, -is_placed)


```


```{r 19,20}

# Build model
tree_model <- rpart( formula = status ~ ., data = train)

# Visualize tree
print(tree_model)

# Plot tree
rpart.plot(tree_model, type = 2, extra = 'auto', fallen.leaves = TRUE)

summary(tree_model)

```

```{r Q21}

# Evaluation

asRules(tree_model)

y_pred <- predict(tree_model, newdata = test, type = 'class')
y_pred

```

```{r Q22}

# Feature importance
feature_importance <- sort(tree_model$variable.importance, decreasing= TRUE)
feature_importance

```

```{r Q23,24}

# rebuild model using rpart.control 
tree_model <- rpart( formula = status ~ ., data = train, control = rpart.control(minsplit = 10, cp = 0.001)) # Change parameters as needed

# Plot tree
rpart.plot(tree_model, type = 2, extra = 'auto', fallen.leaves = TRUE)


```

## Conclusion (questions 23,24)

### Two strong rules

1- ssc_p > 68 and gender = F will give 28% support and prob 1

2- ssc_p < 68 and degree_p > 66 will give  15% support and prob 0.82 

### Pruning parameters

CP: A lower cp will allow the tree to grow larger with more splits, a higher cp will stop the tree earlier.

Minsplit: sets the minimum number of observations that must exist in a node for a split to happen. Lower minsplit leads to a deeper tree with more splits.

Minbucket: the minimum number of observations that a terminal node (leaf) must have. Lower minbucket means more splits.

Maxdepth: Limits the maximum depth of the tree. Higher maxdepth leads to a deeper tree.

